Fractional Types

### 2013-02-23 frac.tex
edits for length, style and precision.  Add a longer paragraph for the groupoid ideas

Technically speaking, obtaining a higher-order extension requires the
construction of a \emph{compact closed category} from the underlying monoidal
category for~{{Pi}}. Although the general idea of such a construction is
well-understood, the details of adapting it to an actual programming language
are subtle (as we explain in the main body of the paper).  Our main novel
technical device to achieving the higher-order extension is a
\emph{fractional type}, which we so named because of its duality with
conventional product types.  By adding a formal (linear) dual to (type
theoretic) products, we can internalize the notion that ``taking an input''
corresponds to negative information, which turns out to be the basis of
first-class functions. The remainder of the paper reviews {{Pi}} and then
introduces the syntax and semantics of the extension with fractional
types. We then study the properties of the extended language and establish
its expressiveness via several constructions and examples that exploit its
ability to model higher-order computations.

@@

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by:
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
\noindent and these are classified by ordinary types:
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b 
\noindent Types include the unit type {{1}}, sum types {{b1 + b2}}, and
products types {{b1*b2}}.  Values includes {{()}} which is the only value of
type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. But these values should
be regarded as largely ancillary.  In particular we do not treat the above
values as first-class citizens.  They only occur when we want to observe the
effect of an isomorphism.

@@

The isomorphisms are extended to form a congruence relation by adding the
following constructors that witness equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
\noindent The syntax is overloaded: we use the same symbol at the value-type level
and at the isomorphism-type level for denoting sums and products.  Hopefully
this will not cause undue confusion.

@@

The extension of the sets of types and values from {{Pi}} to {{PiTF}} is
simple:

%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. We have
however some informal intuition about the \emph{meaning} of such fractional
types and values. (See Sec.~\ref{sec:cat} for more discussion about this
point.) In more detail, given a symmetric monoidal category modeling a
certain language (like {{Pi}}), one can obtain a higher-order extension of
the language by extending the category to a \emph{compact closed} one, i.e.,
a category in which morphisms are representable as objects. The key to this
extension is to add dual objects (in our case the fractionals) satisfying the
following isomorphism witnessed by two new type indexed combinators 
{{eta*_b}} and {{eps*_b}}:

%subcode{bnf} include main
%! columnStyle = r r c l l
%eta*_b :&  1 & <-> & 1/b * b &: eps*_b

%% {{eta*_b : 1 <-> 1/b * b : eps*_b}}

@@

From the categorical perspective, versions of compact closed categories were
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the
generalization of monoidal categories to model quantum computation.  The
first and most intuitive example of such categories is the category of finite
sets and relations. In that case, the dual operation (i.e., the fractional
type) collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
interpretation of fractionals, this is still an interesting category as it
provides a model for a reversible higher-order language. Indeed, any relation
is reversible via the standard \emph{converse} operation on a relation.
Furthermore, any relation between {{b1}} and {{b2}} can be
represented as a subset of {{b1 * b2}}, i.e., as a value. A more interesting
example is the category of finite dimensional vector spaces and linear maps
(or even the category of finite dimensional Hilbert spaces). In such
categories the dual operation takes a vector space (consisting of ``column
vectors'') to the dual vector space (consisting of ``row vectors''). These
vector spaces are isomorphic, but they are nevertheless quite different,
being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words, the
category provides a more refined model in which isomorphic
negative-information values and positive-information are not identified.

In this section, we provide a simple semantics of {{PiTF}} in the category of
finite sets and relations. In other words, we interpret each combinator 
{{c : b1 <-> b2}} as a \emph{relation} between the sets denoted by {{b1}} and
{{b2}} respectively. As we illustrate, this semantics allows \emph{arbitrary}
relations (including the empty relation and relations that are not
isomorphisms) to be represented as values in {{PiTF}}.  In Sec.~\ref{sec:cat}
we discuss more refined semantics that are more suitable for richer
categories.

@@

The denotation of combinators as relations is given below. We specify
relations using a deductive system whose judgments are of the form 
{{ v1 ~~c~~ v2 }} indicating that the pair {{(v1,v2)}} is in the relation 
denoted by {{c}}.

@@

%% XXX: diagrams need unit wires(?)
\begin{multicols}{2}
  \dgm{eta_times2}

@@

In the following, we use {{bool}} as an abbreviation of {{1+1}} 
with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
abbreviation for {{right ()}}. In addition, for the sake of readability, the
circuit diagrams in this section will elide trivial unit wires, such as in
{{eta}} and {{eps}}.

@@

\dgm{compose-times3}

\noindent
We can also derive currying (and dually, uncurrying) combinators. Observe that
the type of {{curry}} needs to be {{b1 * b2 :-* b3 <-> b1 :-* (b2 :-* b3)}},
which can be written in mathematical notation as {{b3/(b1 * b2) = (b3 * 1/b2) *
(1/b1)}}. This means {{curry}} can be written using {{recip}}, the
implementation of the mathematical identity {{1/(b1 * b2) = 1/b1 * 1/b2}}:

%subcode{opsem} include main
%! columnStyle = rcl
% recip &:& 1 / (b1 * b2) <-> 1 / b1 * 1 / b2
% recip &=& (uniti) (;) (uniti) (;) (assocl*) (;) ((eta*_{b1} (*) eta*_{b2}) (*) id)
%     &&    (;) (reorder (*) id) (;) (assocr*) (;) (id (*) swap*) (id (*) eps*)
%     &&    (;) swap* (;) (unite)

@@

%subcode{opsem} include main
%! columnStyle = rcl
% curry &:& b1 (x) b2 :-* b3 <-> b1 :-* (b2 :-* b3)
% curry &=& (id * recip) (;) (id * swap*) (;) assocl*

%%%%%%%%%%%%
\subsection{Duality}

The final remarkable property of fractional types is that they
are involutive.  In particular, we have:

@@

\dgm{involution-times}

@@

Mathematically speaking, recursion and iteration can be expressed using
categorical trace
operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
language like {{Pi}} there are two natural families of trace operators
that can be defined, an additive family (explored in detail in our previous
work~\cite{rc2011}) and a multiplicative family (which is the focus of this
section).

Given a computation {{c:b1 * b2 <-> b1 * b3}},we
want to cancel the common type {{b1}}, to produce a new combinator
{{trace*~c:b2<->b3}}. For the evaluation of such a combinator, we are only
given a value {{v2 : b2}}. But to evaluate {{c}}, we must provide a pair
consisting of ``some'' value {{v1 : b1}} together with the given {{v2 : b2}}.
The needed value {{v1 : b1}} cannot be arbitrary: it must be such that it
is the value produced as the first component of the result.  In general, there
may be several such fixed-point values, or none.  A few examples should help
explain the subtleties.

\begin{example}
Consider the combinator

{{c = id}}

@@

We can use the relational semantics of definition~\ref{def:relational-PiTF}
to calculate the result of applying
{{trace*_{bool} (swap+ (*) id)}} to {{false}}.  We use the usual conversion
of a relation from a subset of {{A x B}} to a function from {{A}} to the
powerset of {{B}} (and Kleisli composition in the powerset monad) for
the computation:
%subcode{opsem} include main
%! columnStyle = ll
% {[ ((),false) ]}  & (uniti)
% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool} * id)
% {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (id (*) not )
% {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
% emptyset & (eps*_{bool} * id)

\noindent  The computation with {{true}} gives the same result.  This
implies that although our semantics is formally reversible, it 
strays from the kind of reversibility that has been achieved with {{Pi}}.
Nevertheless, the next subsection illustrates why this is still useful.

@@

\noindent We can define a combinator {{c_R}} whose denotation is {{R}}.
The key idea is to find a combinator {{cInner : (a * bool) <-> (a * bool)}}
for some type {{a}} such that {{c_R}} == {{trace* cInner}}.
A little experimentation shows that we get the desired behavior 
as follows:
% %subcode{proof} include main
% %@ ~
% %@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
% %----
% %@ ~
% %@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
% %----
% %@ ~
% %@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
% %----
% %@ ~
% %@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
% %----
% %@ ~
% %@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
% %----
% %@ ~
% %@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
% %----
% %@ ~
% %@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
% %----
% %@ ~
% %@@ ((true,true),true) ~~cInner~~ ((false,false),true) 
% In the first three lines, the first argument is a fixed point and hence the
% multiplicative trace would map the second input to the second output
% producing the desired relation. In the remaining five cases, the first
% argument is not a fixed point and hence all these cases would be rejected as
% solutions to the constraint imposed by the multiplicative trace. It simply
% remains to find the actual realization of {{cInner}} that would produce the
% behavior above.

%subcode{opsem} include main
%! columnStyle = rcl
% cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;) 
%        &&  toffoli (;) ((swap* (;) cnot (;) swap*) (*) id) (;)
%        &&  toffoli (;) (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;)
%        &&  toffoli (;) (cnot (*) id))
\noindent where {{toffoli}} and {{cnot}} were defined (for {{Pi}}) in
\cite{rc2011}.
% %     
% % c_R &:& bool <-> bool
% % c_R &=& trace* cInner

@@

For these reasons, it is interesting to consider other possible semantic
interpretations of fractionals. A natural alternative is to use the
``canonical'' compact closed category, that is finite dimensional vector
spaces and linear maps\cite{Selinger:2011:FDH:1942319.1942398,Hasegawa:2008:FDV:1805839.1805859} over fields of characteristic $0$. Let us fix an arbitrary
field $k$ of characteristic $0$.
Such an interpretation has the following features:

\begin{itemize}
\item each type {{b}} in {{PiTF}} is interpreted as a finite dimensional
  vector space $\mathbf{V}_b$ over the field $k$. Of particular
  interest:
  \begin{itemize}
  \item every vector space contains a zero vector which means that the type
    {{0}} (if included in {{PiTF}} would not be the ``empty'' type.
    Furthermore all combinators would be
    strict as they would have to map the zero vector to itself. 
  \item the type {{1}} is interpreted as a 1-dimensional vector space and hence
    is isomorphic to the underlying field.
  \item the fractional type {{1/b}} is interpreted as the dual vector space
    to the vector space $\mathbf{V}_b$ representing {{b}} consisting of all
    the \emph{linear functionals} on $\mathbf{V}_b$.
  \item one can then validate certain desirable properties: {{1/(1/b)}} is
    isomorphic to {{b}}; and {{eps*_b}} corresponds to a bilinear form which
    maps a dual vector and a vector to a field element.
  \end{itemize}
\end{itemize}

Although this semantics appears to have ``better'' properties than the
relational one, we argue that it is not \emph{the} ``perfect'' semantics. By
including the zero vector, the language has
morphisms that do not correspond to isomorphisms and it is difficult to
reconcile the interpretation with the view that the types are rational
numbers. What would really be a ``perfect'' interpretation is one in which we
can only express {{Pi}}-isomorphisms as first class values and in which the
types are interpreted in a way that is consistent with the rational
numbers. Fortunately, there is promising significant work on the groupoid
interpretation of type theory and on the categorification of the rational
numbers that we outline next.

The fundamental idea of both lines of previous work is that it is possible to
build sets (with additional structure) with a \emph{fractional cardinality}.

%% EXPAND on this

@@

Give the denotation of a few combinators. The denotation of {{eta_{1+1} }} is
the relation:

{{ {[ (1/(), (left(), left())), (1/(), (right(),right ())) ]} }}

give some intuition; example with eta/epsilon

>>>>

Technically speaking, obtaining a higher-order extension requires the
construction of a \emph{compact closed category} from the underlying monoidal
category for~{{Pi}}. Although the general idea of such a construction is
well-understood, the details of adapting it to an actual programming language
are subtle.  Our main novel
technical device to achieving the higher-order extension is a
\emph{fractional type}, which we so named because of its duality with
conventional product types.  By adding a formal (linear) dual to (type
theoretic) products, we can internalize the notion that ``taking an input''
corresponds to negative information, which turns out to be the basis of
first-class functions. The remainder of the paper reviews {{Pi}} and then
introduces the syntax and semantics of the extension with fractional
types. We then study the properties of the extended language and establish
its expressiveness via several constructions and examples that exploit its
ability to model higher-order computations.

@@

\paragraph*{Data.}
We view {{Pi}} as having two levels:  it has traditional values, given by:
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
\noindent and these are classified by ordinary types:
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b 
\noindent Types include the unit type {{1}}, sum types {{b1 + b2}}, and
products types {{b1*b2}}.  Values includes {{()}} which is the only value of
type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. But these values should
be regarded as largely ancillary: we do not treat them as first-class citizens,
and they only occur when observing the effect of an isomorphism.

@@

The isomorphisms are extended to form a congruence relation by adding the
following constructors that witness equivalence and compatible closure:

%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4
\noindent The syntax is overloaded: we use the same symbol at the value-type level
and at the isomorphism-type level for denoting sums and products.  Hopefully
this will not cause undue confusion.

@@

The extension of the sets of types and values from {{Pi}} to {{PiTF}} is
simple:
%subcode{bnf} include main
% value types, b ::= ... | 1/b
% values, v ::= ... | 1/v
For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}. We have
however some informal intuition about the \emph{meaning} of such fractional
types and values. (See Sec.~\ref{sec:cat} for more discussion about this
point.) In more detail, given a symmetric monoidal category modeling a
certain language (like {{Pi}}), one can obtain a higher-order extension of
the language by extending the category to a \emph{compact closed} one, i.e.,
a category in which morphisms are representable as objects. The key to this
extension is to add dual objects (in our case the fractionals) satisfying the
following isomorphism witnessed by two new type indexed combinators 
{{eta*_b}} and {{eps*_b}}:
%subcode{bnf} include main
%! columnStyle = r r c l l
%eta*_b :&  1 & <-> & 1/b * b &: eps*_b
%% XXX: diagrams need unit wires(?)
\begin{multicols}{2}
  \dgm{eta_times2}

@@

From the categorical perspective, versions of compact closed categories were
introduced by Abramsky and Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878}
and by Selinger~\cite{Selinger:2007:DCC:1229185.1229207} as the
generalization of monoidal categories to model quantum computation.  The
first and most intuitive example of such categories is the category of finite
sets and relations. In that case, the dual operation (i.e., the fractional
type) collapses by defining {{1/b}} to be {{b}}. In spite of this degenerate
interpretation of fractionals, this is still an interesting category as it
provides a model for a reversible higher-order language. Indeed, any relation
is reversible via the standard \emph{converse} operation on a relation.
Furthermore, as any relation between {{b1}} and {{b2}} can be
represented as a subset of {{b1 * b2}}, i.e., as a value. A more interesting
example is the category of finite dimensional vector spaces and linear maps
(or even the category of finite dimensional Hilbert spaces). In such
categories the dual operation takes a vector space (consisting of ``column
vectors'') to the dual vector space (consisting of ``row vectors''). While
these are isomorphic, but they are nevertheless quite different,
being $(1,0)$ and $(0,1)$ tensors (respectively).  In other words, the
category provides a more refined model in which isomorphic
negative-information values and positive-information are not identified.

In this section, we provide a simple semantics of {{PiTF}} in the category of
finite sets and relations. We interpret each combinator 
{{c : b1 <-> b2}} as a \emph{relation} between the sets denoted by {{b1}} and
{{b2}} respectively. As we illustrate, this semantics allows \emph{arbitrary}
relations (including the empty relation and relations that are not
isomorphisms) to be represented as values in {{PiTF}}.  In Sec.~\ref{sec:cat}
we discuss more refined semantics that are more suitable for richer
categories.

@@

We specify
relations using a deductive system whose judgments are of the form 
{{ v1 ~~c~~ v2 }} indicating that the pair {{(v1,v2)}} is in the relation 
denoted by {{c}}.

@@

@@

In the following, we use {{bool}} as an abbreviation of {{1+1}} 
with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
abbreviation for {{right ()}}. In addition, for the sake of readability, the
circuit diagrams in this section will elide trivial unit wires, such as in
{{eta}} and {{epsilon}}.

@@

% \dgm{compose-times3}

\noindent
We can also derive currying (and dually, uncurrying) combinators. Observe that
the type of {{curry}} needs to be {{(b1 * b2 :-* b3) <-> (b1 :-* (b2 :-* b3))}},
which can be written in mathematical notation as 
{{1/(b1 * b2) * b3 = (1/b1) * ((1/b1) * b3)}}. This means {{curry}} can be
written using {{recip}}, the implementation of the mathematical identity
{{1/(b1 * b2) = 1/b1 * 1/b2}}:

%subcode{opsem} include main
%! columnStyle = rcl
% recip &:& 1 / (b1 * b2) <-> 1 / b1 * 1 / b2
% recip &=& (uniti) (;) (uniti) (;) (assocl*) (;) ((eta*_{b1} (*) eta*_{b2}) (*) id) (;)
%     && (reorder (*) id) (;) (assocr*) (;) (id (*) swap*) (id (*) eps*) (;) swap* (;) (unite)

@@

%subcode{opsem} include main
%! columnStyle = rcl
% curry &:& b1 * b2 :-* b3 <-> b1 :-* (b2 :-* b3)
% curry &=& (id * recip) (;) (id * swap*) (;) assocl*
\noindent That {{recip}} is the heart of currying seems quite remarkable.
%%%%%%%%%%%%
\subsection{Duality}

The final remarkable property of fractional types is that they
are involutive:

@@

% jacques: elide for now, to save space, as this diagram is backwards.
% \dgm{involution-times}

@@

Mathematically speaking, recursion and iteration can be expressed using
categorical trace
operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
language like {{Pi}} there are two natural families of trace operators
that can be defined, an additive family (explored in detail in our previous
work~\cite{rc2011}) and a multiplicative family, which we will focus on now.

Given a computation {{c:b1 * b2 <-> b1 * b3}}, we
want to cancel the common type {{b1}}, to produce a new combinator
{{trace*~c:b2<->b3}}. For its evaluation, we are only
given a value {{v2 : b2}}. But to evaluate {{c}}, we must provide a pair
consisting of ``some'' value {{v1 : b1}} together with the given {{v2 : b2}}.
The needed value {{v1 : b1}} cannot be arbitrary: it must be such that it
is the value produced as the first component of the result.  In general, there
may be several such fixed-point values, or none.  A few examples should help
explain the subtleties.

\begin{example}
Consider the combinator {{c : bool * 1 <-> bool * 1}}, with {{c = id}}.

@@

We can use the relational semantics of definition~\ref{def:relational-PiTF}
to calculate the result of applying
{{trace*_{bool} (swap+ (*) id)}} to {{false}}.  We use the usual conversion
of a relation from a subset of {{A * B}} to a function from {{A}} to the
powerset of {{B}} (and Kleisli composition in the powerset monad) for
the computation:
%subcode{opsem} include main
%! columnStyle = ll
% {[ ((),false) ]}  & (uniti)
% {[ ((1/false,false),false), ((1/true,true),false) ]} & (eta*_{bool} * id)
% {[ (1/false, (false,false)), (1/true,(true,false)) ]} & (assocr*)
% {[ (1/false, (true,false)), (1/true,(false,false)) ]} & (id (*) not )
% {[ ((1/false,true),false), ((1/true,false),false) ]} & (assocl*)
% emptyset & (eps*_{bool} * id)

\noindent  The computation with {{true}} gives the same result.  This
implies that although our semantics is formally reversible, it 
does not result in isomorphisms.
Nevertheless, the next subsection illustrates why this is still useful.

@@

\noindent We can define a combinator {{c_R}} whose denotation is {{R}},
by defining a combinator {{cInner : (a * bool) <-> (a * bool)}}
for some type {{a}} such that {{c_R = trace* cInner}}:

% %subcode{proof} include main
% %@ ~
% %@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
% %----
% %@ ~
% %@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
% %----
% %@ ~
% %@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
% %----
% %@ ~
% %@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
% %----
% %@ ~
% %@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
% %----
% %@ ~
% %@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
% %----
% %@ ~
% %@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
% %----
% %@ ~
% %@@ ((true,true),true) ~~cInner~~ ((false,false),true) 
% In the first three lines, the first argument is a fixed point and hence the
% multiplicative trace would map the second input to the second output
% producing the desired relation. In the remaining five cases, the first
% argument is not a fixed point and hence all these cases would be rejected as
% solutions to the constraint imposed by the multiplicative trace. It simply
% remains to find the actual realization of {{cInner}} that would produce the
% behavior above.

%subcode{opsem} include main
%! columnStyle = rcl
% cInner &:& (bool * bool) * bool <-> (bool * bool) * bool
% cInner &=& sym (swap* (;) assocl* (;) (cnot (*) id) (;) assocr* (;) swap* (;)  toffoli (;)
%        &&  ((swap* (;) cnot (;) swap*) (*) id) (;) toffoli (;)
%        &&  (assocr* (;) swap* (;) toffoli (;) swap* (;) assocl*) (;) toffoli (;) (cnot (*) id))

\noindent where {{toffoli}} and {{cnot}} were defined for {{Pi}} in
\cite{rc2011}.
% %     
% % c_R &:& bool <-> bool
% % c_R &=& trace* cInner

@@

For these reasons, it is interesting to consider other possible semantic
interpretations of fractionals. A natural alternative is to use the
``canonical'' compact closed category, that is finite dimensional vector
spaces and linear maps\cite{Selinger:2011:FDH:1942319.1942398,Hasegawa:2008:FDV:1805839.1805859} over fields of characteristic $0$. Let us fix an arbitrary
field $k$ of characteristic $0$.
Then each type {{b}} in {{PiTF}} is interpreted as a finite dimensional
  vector space $\mathbf{V}_b$ over the field $k$. In particular:
\begin{itemize}
  \item every vector space contains a zero vector which means that the type
    {{0}} (if included in {{PiTF}}) would not be the ``empty'' type.
    Furthermore all combinators would be
    strict as they would have to map the zero vector to itself. 
  \item the type {{1}} is interpreted as a 1-dimensional vector space and hence
    is isomorphic to the underlying field.
  \item the fractional type {{1/b}} is interpreted as the dual vector space
    to the vector space $\mathbf{V}_b$ representing {{b}} consisting of all
    the \emph{linear functionals} on $\mathbf{V}_b$.
  \item one can then validate certain desirable properties: {{1/(1/b)}} is
    isomorphic to {{b}}; and {{eps*_b}} corresponds to a bilinear form which
    maps a dual vector and a vector to a field element.
\end{itemize}

Although this semantics appears to have ``better'' properties than the
relational one, we argue that it is not \emph{the} ``perfect'' semantics. By
including the zero vector, the language has
morphisms that do not correspond to isomorphisms (in particular it allows
partial morphisms by treating the {{0}} element of the zero-dimensional 
vector space as a canonical ``undefined'' value).  It is also difficult to
reconcile the interpretation with the view that the types correspond
to the (positive) rational numbers, something we are actively seeking. What 
would really be a ``perfect'' interpretation is one in which we
can only express {{Pi}}-isomorphisms as first class values and in which the
types are interpreted in a way that is consistent with the rational
numbers.

Fortunately, there is promising significant work on the groupoid
interpretation of type theory~\cite{Hofmann96thegroupoid} and on the 
categorification of the rational numbers~\cite{math/9802029} that may well give
us the model we desire.  The fundamental idea in both cases is that groupoids
(viewed as sets with explicit isomorphisms as morphisms) naturally have a
\emph{fractional cardinality}.  Types would be interpreted as groupoids, and
terms would be (invertible) groupoid actions.  The remaining challenge is to
identify those groupoid actions which are proper generalizations of
isomorphisms \emph{and} can be represented as groupoids, so as to obtain a
proper interpretation for a higher-order language.  As the category of
groupoids is cartesian closed, this appears eminently feasible.

@@
